Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features

### 2021-04-28 publication/manuscript.Rmd
switch to springer template

---
title: Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features
author:
  - name: Florian Pargent
    email: florian.pargent@psy.lmu.de
    affiliation: LMU Munich 1
    footnote: 1
  - name: Florian Pfisterer
    #email: florian.pfisterer@stat.uni-muenchen.de
    affiliation: LMU Munich 2
  - name: Janek Thomas
    #email: janek.thomas@iis.fraunhofer.de
    affiliation: Fraunhofer
  - name: Bernd Bischl
    #email: bernd.bischl@stat.uni-muenchen.de
    affiliation: LMU Munich 2
address:
  - code: LMU Munich 1
    address: Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstraße 13, 80802 Munich, Germany
  - code: LMU Munich 2
    address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany
  - code: Fraunhofer
    address: Fraunhofer Institute for Integrated Circuits, Hansastraße 32, 8068 Munich, Germany
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Because most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect during data analysis. An often encountered problem are high cardinality features, i.e. unordered categorical predictor variables with a high number of levels. We study techniques that yield numeric representations of categorical variables which can then be used in subsequent ML applications. We focus on the impact of those techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique. We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbours, support vector machine) using datasets from regression, binary- and multiclass- classification settings. Throughout our study, regularized versions of target encoding (i.e. using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results. Traditional encodings that make unreasonable assumptions to map levels to integers (e.g. integer encoding) or to reduce the number of levels (possibly based on target information, e.g. leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
keywords: supervised machine learning, benchmark, high-cardinality categorical features, target encoding, impact encoding, dummy encoding

journal: "Computational Statistics and Data Analysis"
date: "`r Sys.Date()`"
bibliography: literature.bib
#linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
output:
  #rticles::elsevier_article: default
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
#output: rticles::elsevier_article

header-includes:
   - \usepackage{algorithm}
   - \usepackage{array}
   - \usepackage[noend]{algpseudocode}
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{amsfonts}
   - \usepackage{mathtools}
   - \usepackage{bm}
   - \usepackage{tikz}
   - \usepackage{threeparttable}
   - \usetikzlibrary{trees,arrows, chains, fit, positioning, calc, shapes, shadows}
   - \setlength\parindent{0pt}
   - \usepackage{pdflscape}
---

>>>>

---
title: Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features
titlerunning: Encoding high cardinality features in supervised ML
authors:
  - name: Florian Pargent \orcidlink{0000-0002-2388-553X}
    email: florian.pargent@psy.lmu.de
    address: Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstraße 13, 80802 Munich, Germany
    footnote: 1
  - name: Florian Pfisterer
    address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany
  - name: Janek Thomas \orcidlink{0000-0003-4511-6245}
    address: Fraunhofer Institute for Integrated Circuits, Hansastraße 32, 8068 Munich, Germany
  - name: Bernd Bischl \orcidlink{0000-0001-6002-6980}
    address: Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany

abstract: |
  Because most machine learning (ML) algorithms are designed for numerical inputs, efficiently encoding categorical variables is a crucial aspect during data analysis. An often encountered problem are high cardinality features, i.e. unordered categorical predictor variables with a high number of levels. We study techniques that yield numeric representations of categorical variables which can then be used in subsequent ML applications. We focus on the impact of those techniques on a subsequent algorithm's predictive performance, and -- if possible -- derive best practices on when to use which technique. We conducted a large-scale benchmark experiment, where we compared different encoding strategies together with five ML algorithms (lasso, random forest, gradient boosting, k-nearest neighbours, support vector machine) using datasets from regression, binary- and multiclass- classification settings. Throughout our study, regularized versions of target encoding (i.e. using target predictions based on the feature levels in the training set as a new numerical feature) consistently provided the best results. Traditional encodings that make unreasonable assumptions to map levels to integers (e.g. integer encoding) or to reduce the number of levels (possibly based on target information, e.g. leaf encoding) before creating binary indicator variables (one-hot or dummy encoding) were not as effective.
keywords:
- supervised machine learning
- benchmark
- high-cardinality categorical features
- target encoding
- impact encoding
- dummy encoding

#date: "`r Sys.Date()`"
bibliography: literature.bib
#linenumbers: true
numbersections: true
output:
  #rticles::elsevier_article: default
  bookdown::pdf_book:
    base_format: rticles::springer_article
bibstyle: spphys

header-includes:
   - \usepackage{algorithm}
   - \usepackage{array}
   - \usepackage[noend]{algpseudocode}
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{amsfonts}
   - \usepackage{mathtools}
   - \usepackage{bm}
   - \usepackage{tikz}
   - \usepackage{threeparttable}
   - \usetikzlibrary{trees,arrows, chains, fit, positioning, calc, shapes, shadows}
   - \setlength\parindent{0pt}
   - \usepackage{pdflscape}
   - \usepackage[T1]{fontenc}
   - \usepackage{orcidlink}
---
